---
title: "Data Visualization - Part 1"
subtitle: "The Good, The Bad and The Ugly"
author: "Scott Stoltzman"
date: "March 14, 2017"
output: html_document
---

```{r setup, results='hide', warning=FALSE, message=FALSE,echo=FALSE}
library(png)
library(grid)
```
----  

# A Lesson on Data Visualization - Choosing the Type  

```{r, fig.align='center',echo=FALSE}
img = readPNG("images/title_photo.png")
grid.raster(img)
```

### Know what you are trying to achieve before starting



Let's load up some libraries and get started.

```{r libraryPrep, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(scales)
```

## Decide on what you're trying to accomplish first.  

Ask yourself the following questions to help drive your decision:  

- Are you making a comparison?
- Are you finding a relationship?
- Are you showing a distribution?
- Are you finding a trend over time?
- Are you showing composition?
  
Once you know which question you are asking, it will keep your mind focused on the outcome and will quickly narrow down your charting options.

#### Rule of Thumb  

- **Trend: ** Column, Line  
- **Comparison: ** Area, Bar, Bullet, Column, Line, Scatter  
- **Relationship: ** Line, Scatter  
- **Distribution: ** Bar, Boxplot, Column  
- **Composition: ** Donut, Pie, Stacked Bar, Stacked Column  
  
Obviously, there are choices beyond these and you need to make your choice wisely. 

Side Note: I ***hate*** donut and pie charts! When used properly, they're terriffic! However, I'm very used to gagging every time one appears on a projector screen due to how frequently they're used inappropriately.

For this project, I'll use some oil production data that I found while digging through http://data.world (pretty great site). The data can be found [here](http://www.eia.gov/dnav/pet/pet_crd_crpdn_adc_mbbl_m.htm)  


```{r dataLoading, results='hide', warning=FALSE, message=FALSE}
#Custom data preparation
#GitHub (linked to at bottom of this post)
source('data_preparation.R')
data = getData()
```

```{r}
head(data)
```

----  

## Trend - Line Chart

**Objective:** Visualize a trend in oil production in the US from 1981 - 2016 by year. I want to illustrate the changes over the time period. This is a very high-level view and only shows us a decline followed by a ramp up at the end of the period. This is ***exploratory visualization.***

I decided to use a line chart to show the trend over time. When using discrete data you should use a column chart to avoid any confusion that in between these years the data actually was simply linear. However, it paints a much clearer picture this way and is not misleading.

#### Poor Version  
The x-axis is a disaster and the y-axis isn't formatted well. While it gets the point across, it's still worthless.


```{r,fig.align='center', fig.width=4}
df = data %>% 
  group_by(Year) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel))

p = ggplot(df,aes(x=Year,y=ThousandBarrel,group=1)) 
p + geom_line(stat='identity') + 
  ggtitle('Oil Production Over Time') + 
  theme(plot.title = element_text(hjust = 0.5),plot.subtitle = element_text(hjust = 0.5)) + 
  xlab('') + ylab('')
```

#### Better Version  
The title gives us a much better understanding of what we're looking at. The chart is slightly wider and the axes are formatted to be legible.

```{r,fig.align='center', fig.width=12}
p = ggplot(df,aes(x=Year,y=ThousandBarrel,group=1)) 
p + geom_line(stat='identity') + 
  ggtitle('Thousand Barrel Oil Production By Year in the U.S.') +
  theme(plot.title = element_text(hjust = 0.5),plot.subtitle = element_text(hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  scale_y_continuous(labels = comma)
```


----

## Comparison - Line Chart  

**Objective**: Identify which states affected the trend the most. Evaluate them simultaneously in order to paint the picture and compare their trends over the time period. From this visual you can see the top states are Alaska, California, Louisiana, Oklahoma, Texas and Wyoming. Texas seems to break the mold quite drastically and drove the spike which occurred after 2010. This is another example of ***exploratory visualization.***

#### Poor Version  
There are far too many colors going on here. Everything at the bottom of the chart is relatively useless and takes our focus away from the big players. 

```{r,warning=FALSE,fig.width=10,message=FALSE}
df = data %>%
  group_by(Location, Year) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel))

df$Year = as.numeric(df$Year)

p = ggplot(df,aes(x=Year,y=ThousandBarrel,col=Location))
p + geom_line(stat='identity') + 
  ggtitle(paste('Oil Production By Year By State in the U.S.')) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

#### Better Version  
This focuses attention on the top producing states. It compares them to each other and shows the trend per state as well.

```{r,warning=FALSE,fig.width=10,message=FALSE}
n=6 #Arbitrary at first, after trying a few, this made the most sense
topN = data %>%
  group_by(Location) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel)) %>%
  arrange(-ThousandBarrel) %>%
  top_n(n)

df = data %>%
  filter(Location %in% topN$Location) %>%
  group_by(Year,Location) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel))

df$Year = as.numeric(df$Year)
df$Location = as.factor(df$Location)

p = ggplot(df,aes(x=Year,y=ThousandBarrel,group=1))
p + geom_line(stat='identity') + 
  ggtitle(paste('Top',as.character(n),'States - Oil Production By Year in the U.S.')) + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  facet_wrap(~Location) + 
  scale_y_continuous(labels = comma) 

```

----

## Relationship - Scatter Plot

**Objective**: See if  data from Alaska and California is correlated (This probably isn't important but it allows us to use the same data).

### Which of these views would you rather see?

#### Poor Version  
Lots of completely irrelevant data! Size of the point should have nothing to do with the year.  

```{r,warning=FALSE,fig.width=10,message=FALSE}
statesList = c('Alaska','California')
df = data %>%
  filter(Location %in% statesList) %>%
  spread(Location,ThousandBarrel) %>%
  select(Alaska,California,Month,Year)

p = ggplot(df,aes(x=Alaska,y=California,col=Month,size=Year))
p + geom_point() + 
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma) +
  ggtitle('Oil Production - CA vs. AK') + 
  theme(plot.title = element_text(hjust = 0.5))

```

#### Better Version  
The trend line is nice because it helps to visualize the relationship even more. While it can sometimes be misleading, it makes sense with our current data.  

```{r,warning=FALSE,fig.width=10,message=FALSE}
df = data %>%
  filter(Location %in% statesList) %>%
  spread(Location,ThousandBarrel) %>%
  select(Alaska,California,Year)

p = ggplot(df,aes(x=Alaska,y=California))
p + geom_point() + 
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = comma) +
  ggtitle('Monthly Thousand Barrel Oil Production 1981-2016 CA vs. AK') + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_smooth(method='lm')

```




## Distribution - Boxplot 

**Objective**: Examine the range of production by state and year over the time period to give us an idea of the variance.

### Which of these views would you rather see?

#### Poor Version  


```{r,warning=FALSE,fig.width=10,message=FALSE}
df = data %>%
  group_by(Year,Location) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel))

p = ggplot(df,aes(x=Location,y=ThousandBarrel))
p + geom_boxplot() + 
  ggtitle('Distribution of Oil Production by State')

```


#### Better Version  
This gives a nice ranking to the plot while still showing their distributions. While it was semi-apparent in the line charts, the variance of Texas is huge compared to the others! We could take this a step further and separate out the big players from the smaller players.

```{r,warning=FALSE,fig.width=10,message=FALSE}
p = ggplot(df,aes(x=reorder(Location,ThousandBarrel),y=ThousandBarrel))
p + geom_boxplot() + 
  scale_y_continuous(labels = comma) +
  ggtitle('Distribution of Annual Oil Production By State (1981 - 2016)') + 
  coord_flip()
```


## Composition - Stacked Bar 

**Objective**: Check out the composition of total production by state.

### Which of these views would you rather see?

#### Poor Version  
My favorite, the beautiful pie chart! There's nothing better than this...

```{r,warning=FALSE,fig.width=10,message=FALSE}
df = data %>%
  group_by(Location) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel)) %>%
  mutate(ThousandBarrel = ThousandBarrel/sum(ThousandBarrel))

df$ThousandBarrel = round(100*df$ThousandBarrel,0)

library(plotrix)
pie(x=df$ThousandBarrel,labels=df$Location,explode=0.1,col=rainbow(nrow(df)),main='Percentage of Oil Production by State')

```


#### Better Version  
The 1980's and 2010's will be missing years in terms of a "decade" due to the data provided (and it's only 2017). While the percentage labels are slightly off center, it's certainly much better than the pie chart. It's not quite "apples-to-apples" for a comparison because I created different decades, but you get the idea.

I also created an "Other" category in order to simplify the output. When you are doing comparisons, it's typically a good idea to find a way to reduce the number of variables in the output while not removing data by dropping it completely.

```{r,warning=FALSE,fig.width=10,message=FALSE}
data$Decade = '1980s'
data$Decade[data$Year >= 1990] = '1990s'
data$Decade[data$Year >= 2000] = '2000s'
data$Decade[data$Year >= 2010] = '2010s'
data$Decade = as.factor(data$Decade)

top5 = data %>%
  group_by(Location) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel)) %>%
  arrange(-ThousandBarrel) %>%
  top_n(5) %>%
  select(Location)

top5List = top5$Location

data$State = "Other"

for(i in 1:length(top5List)){
  data$State[data$Location == top5List[i]] = top5List[i]
}

df = data %>%
  group_by(Decade,State) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel)) %>%
  mutate(ThousandBarrel = ThousandBarrel/sum(ThousandBarrel))

df$ThousandBarrel = round(df$ThousandBarrel,3)
df$text = paste(round(100*df$ThousandBarrel,0),'%', sep='')

p = ggplot(df,aes(x=Decade,y=ThousandBarrel,col=reorder(State,ThousandBarrel),fill=reorder(State,ThousandBarrel)))
p + geom_bar(stat='identity') + 
  geom_text(aes(label=text),col='Black',size = 4, hjust = 0.5, vjust = 3, position = "stack") + 
  scale_y_continuous(labels = percent) +
  ggtitle('Percentage of Top Oil Producing States by Decade') + 
  guides(fill=guide_legend(title='State'),col=guide_legend(title='State')) + 
  theme(plot.title = element_text(hjust = 0.5))

```




### Some other fun concepts are below!  
Some of them are nice, others are terrible! I won't comment on any of them, but I felt it was necessary to include some other ideas I toyed around with. 

Have fun with your data visualizations. The charts I showed here are extremely simple. Being creative by using things other than R wind up making visuals people can remember. There are plenty of examples around, but they all tend to follow basic principles of design. There are ***A TON*** of good books out there on this topic. 

Now it's your turn!


```{r}
df = data %>% 
  group_by(Location) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel)) %>%
  arrange(-ThousandBarrel)
p = ggplot(df,aes(x=reorder(Location,ThousandBarrel),y=ThousandBarrel))
p + geom_bar(stat='identity') + 
  ggtitle('Oil Production 1981 - 2016 By Location') + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  coord_flip()
```





```{r}
top10 = data %>%
  group_by(Location) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel)) %>%
  arrange(-ThousandBarrel) %>%
  top_n(10)
print(top10)

df = data %>% 
  group_by(Location,Year) %>%
  filter(Location %in% top10$Location) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel)) 
p = ggplot(df,aes(x=Year,y=ThousandBarrel,col=Location,fill=Location))
p + geom_bar(stat='identity') + 
  ggtitle('Oil Production - Top 10 States') + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```



```{r, fig.width=10}
df = data %>%
  filter(Year == 1990)%>%
  group_by(Location) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel))
df$Location = tolower(df$Location)

#Add States without data
States = data.frame(Location = tolower(as.character(state.name)))
missingStates = States$Location[!(States$Location %in% df$Location)]
appendData = data.frame(Location=missingStates,ThousandBarrel=0)
df = rbind(df,appendData)

states_map <- map_data("state")

ggplot(df, aes(map_id = Location)) + 
    geom_map(aes(fill=ThousandBarrel), map = states_map) +
    expand_limits(x = states_map$long, y = states_map$lat)

```


```{r}
df = data %>% 
  filter(Location == 'Texas') %>%
  group_by(Year,Month) %>%
  summarise(ThousandBarrel = sum(ThousandBarrel))

p = ggplot(df,aes(x=Month,y=ThousandBarrel))
p + geom_line(stat='identity',aes(group=Year,col=Year)) + 
  ggtitle('Oil Production By Year in the U.S.') + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


As always, the code used in this post is on my [GitHub](https://github.com/stoltzmaniac/Data-Visualization-Lesson)